---
title: "Classifying the Perceived Health of New York City Street Trees"
author: "STAT5003 W12G04"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    self-contained: true
    number-sections: true
    include-in-header:
      text: |
        <style>
        @media print {
          @page { size: A4; margin: 12mm; }
          body { font-size: 10pt; line-height: 1.25; }
          h1 { font-size: 18pt; margin-top: 0.7rem; margin-bottom: 0.4rem; }
          h2 { font-size: 14pt; margin-top: 0.6rem; margin-bottom: 0.3rem; }
          p { margin-bottom: 0.4rem; }
          table { font-size: 9pt; }
          figure {
            break-inside: avoid;
            margin-top: 0.35rem;
            margin-bottom: 0.45rem;
          }
        }
        </style>
execute:
  warning: false
  message: false
---

<!-- Replace all square-bracketed placeholders before submission. -->

# Framing & Dataset Selection

## Research question and motivation

**Research question:** *To what extent can observable tree characteristics,
site conditions, and geographic information classify the perceived health of
living New York City street trees as Good, Fair, or Poor?*

This is a multi-class classification problem with practical relevance to the
New York City Department of Parks and Recreation (NYC Parks), local councils,
community tree stewards, and residents. A reliable model could support the
prioritisation of professional inspections and maintenance when it is not
feasible to inspect every tree simultaneously. It should be treated as a
screening tool rather than an arboricultural diagnosis because the outcome is
the census worker's *perception* of health, not a laboratory measurement (New
York City Department of Parks and Recreation [NYC Parks], 2016). The analysis
is predictive and observational; associations must not be interpreted as
causal effects.

```{r}
#| label: setup
#| results: hide

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(scales)
library(patchwork)
library(knitr)

data_file <- "2015_Street_Tree_Census_-_Tree_Data_20260912.csv"

trees_raw <- read_csv(
  data_file,
  na = c("", "NA", "N/A", "NULL"),
  show_col_types = FALSE,
  progress = FALSE
)

issue_vars <- c(
  "root_stone", "root_grate", "root_other",
  "trunk_wire", "trnk_light", "trnk_other",
  "brch_light", "brch_shoe", "brch_other"
)

trees <- trees_raw |>
  filter(status == "Alive", health %in% c("Good", "Fair", "Poor")) |>
  mutate(
    health = factor(health, levels = c("Good", "Fair", "Poor")),
    problem_count = rowSums(
      across(all_of(issue_vars), ~ .x == "Yes"),
      na.rm = TRUE
    ),
    problem_group = factor(
      if_else(problem_count >= 3, "3+", as.character(problem_count)),
      levels = c("0", "1", "2", "3+")
    )
  )

health_cols <- c(Good = "#2E7D32", Fair = "#F9A825", Poor = "#C62828")

theme_set(
  theme_minimal(base_size = 10) +
    theme(
      plot.title = element_text(face = "bold", size = 11),
      plot.subtitle = element_text(size = 9),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
)
```

# Consider & Gather Data

## Source and scope

The data are the **2015 Street Tree Census--Tree Data**, collected by
volunteers and staff coordinated by NYC Parks and partner organisations. The
census records species, diameter, site conditions, perceived health, and
location for individual street-tree points. Data collection ran from May 2015
to October 2016, so the data represent a historical snapshot rather than the
current condition of New York City's trees (NYC Parks, 2024). The dataset and
data dictionary were obtained from the official
[NYC Open Data portal](https://data.cityofnewyork.us/Environment/2015-Street-Tree-Census-Tree-Data/uvpi-gqnh),
not Kaggle, an R built-in dataset, or course material.

Health is undefined for dead trees and stumps. We therefore restrict the
classification cohort to living trees with a recorded Good, Fair, or Poor
health label. This leaves 652,172 observations; only one living record is
excluded because its health value is missing.

## Suitability against the project requirements

```{r}
#| label: dataset-requirements

requirement_table <- tibble(
  Requirement = c("Large", "Messy", "Multi-class"),
  `Evidence in this dataset` = c(
    paste0(comma(nrow(trees_raw)), " rows and ", ncol(trees_raw),
           " columns; the CSV is approximately 285 MB."),
    paste0(
      comma(sum(is.na(trees_raw$health))),
      " missing health values, structurally concentrated among dead trees ",
      "and stumps; diameter also contains values requiring review."
    ),
    paste0(
      "Good/Fair/Poor outcome with class shares of ",
      percent(mean(trees$health == "Good"), accuracy = 0.1), ", ",
      percent(mean(trees$health == "Fair"), accuracy = 0.1), " and ",
      percent(mean(trees$health == "Poor"), accuracy = 0.1), "."
    )
  )
)

kable(requirement_table, caption = "Evidence that the dataset meets three project requirements")
```

The dataset is not claimed to be **Complex** under the assignment definition
because it has 45 rather than more than 100 features, and it is not claimed to
be **Integrated** because no external data source has been joined. For
computational efficiency, the 285 MB file is loaded once, summaries are
vectorised, and later tree-based model tuning will use parallel processing and
efficient implementations where available.

# Exploratory Data Analysis

## Missingness and outcome distribution

```{r}
#| label: fig-missing-class
#| fig-width: 12
#| fig-height: 4.2
#| fig-cap: "Structural missingness and health-class imbalance."

missing_vars <- c(
  "health", "spc_common", "steward", "guards",
  "sidewalk", "problems", "bin", "bbl"
)

missing_by_status <- trees_raw |>
  group_by(status) |>
  summarise(across(all_of(missing_vars), ~ mean(is.na(.x))), .groups = "drop") |>
  pivot_longer(-status, names_to = "variable", values_to = "missing_rate") |>
  mutate(status = factor(status, levels = c("Alive", "Dead", "Stump")))

p_missing <- ggplot(
  missing_by_status,
  aes(x = status, y = fct_rev(variable), fill = missing_rate)
) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(
    aes(
      label = percent(missing_rate, accuracy = 0.1),
      colour = missing_rate > 0.5
    ),
    size = 2.7,
    show.legend = FALSE
  ) +
  scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "black")) +
  scale_fill_gradient(
    low = "#F7FBFF", high = "#08519C", labels = label_percent()
  ) +
  labs(title = "Missingness depends on tree status", x = NULL, y = NULL,
       fill = "Missing")

class_dist <- trees |>
  count(health, name = "n") |>
  mutate(proportion = n / sum(n))

p_class <- ggplot(class_dist, aes(health, n, fill = health)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(
    aes(label = paste0(comma(n), "\n", percent(proportion, accuracy = 0.1))),
    vjust = -0.15,
    size = 3
  ) +
  scale_fill_manual(values = health_cols) +
  scale_y_continuous(
    labels = label_comma(), expand = expansion(mult = c(0, 0.14))
  ) +
  labs(title = "Living-tree health is imbalanced", x = "Health", y = "Trees")

(p_missing | p_class) +
  plot_layout(widths = c(1.2, 1)) +
  plot_annotation(tag_levels = "A")
```

The condition-related fields are almost entirely absent for Dead and Stump
records because they were not applicable, rather than because of random data
loss. After defining the living-tree cohort, the outcome remains strongly
imbalanced: Good accounts for 81.1%, Fair 14.8%, and Poor only 4.1%. A trivial
always-Good classifier would therefore achieve 81.1% accuracy, demonstrating
why accuracy alone is unsuitable.

## Univariate distributions and outliers

```{r}
#| label: fig-univariate
#| fig-width: 12
#| fig-height: 4.2
#| fig-cap: "Distributions of diameter and common species. Extreme diameters are retained in the audit but omitted from the displayed scale."

top_species <- trees |>
  count(spc_common, sort = TRUE, name = "n") |>
  slice_head(n = 12)

p_dbh <- ggplot(trees, aes(health, tree_dbh, fill = health)) +
  geom_boxplot(outlier.shape = NA, width = 0.65) +
  coord_cartesian(ylim = c(0, quantile(trees$tree_dbh, 0.999, na.rm = TRUE))) +
  scale_fill_manual(values = health_cols, guide = "none") +
  labs(title = "Diameter varies across health classes",
       subtitle = "Display ends at the 99.9th percentile",
       x = "Health", y = "Diameter at breast height (inches)")

p_species_count <- top_species |>
  mutate(spc_common = fct_reorder(spc_common, n)) |>
  ggplot(aes(n, spc_common)) +
  geom_col(fill = "#4C956C", width = 0.7) +
  scale_x_continuous(labels = label_comma()) +
  labs(title = "Most common living-tree species", x = "Trees", y = NULL)

(p_dbh | p_species_count) +
  plot_layout(widths = c(0.9, 1.1)) +
  plot_annotation(tag_levels = "A")
```

Median diameter is 10 inches (mean 11.7; 99th percentile 37), but the maximum
is 425 inches. There are 222 living trees recorded with zero diameter and 67
above 100 inches. These observations require plausibility checks and
sensitivity analysis; they should not be deleted solely because an automated
outlier rule identifies them. Species is also concentrated: London planetree
alone represents 87,014 living trees (13.3%).

## Feature--outcome relationships

```{r}
#| label: fig-feature-outcome
#| fig-width: 12
#| fig-height: 4.2
#| fig-cap: "Health composition by recorded problem burden and observed Poor-health share by species."

health_by_problem <- trees |>
  count(problem_group, health, .drop = FALSE, name = "n") |>
  group_by(problem_group) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()

p_problem <- ggplot(health_by_problem, aes(problem_group, proportion, fill = health)) +
  geom_col() +
  scale_fill_manual(values = health_cols) +
  scale_y_continuous(labels = label_percent()) +
  labs(title = "Health composition changes with recorded problem burden",
       x = "Number of recorded problems", y = "Share of trees", fill = "Health")

poor_species <- trees |>
  filter(!is.na(spc_common)) |>
  group_by(spc_common) |>
  summarise(n = n(), poor_share = mean(health == "Poor"), .groups = "drop") |>
  filter(n >= 1000) |>
  slice_max(poor_share, n = 12) |>
  mutate(spc_common = fct_reorder(spc_common, poor_share))

p_species_health <- ggplot(poor_species, aes(poor_share, spc_common)) +
  geom_col(fill = "#B85C5C", width = 0.7) +
  scale_x_continuous(labels = label_percent()) +
  labs(title = "Poor-health share varies among common species",
       subtitle = "Species shown have at least 1,000 living-tree records",
       x = "Poor-health share", y = NULL)

(p_problem | p_species_health) +
  plot_layout(widths = c(1, 1.1)) +
  plot_annotation(tag_levels = "A")
```

The exploratory relationships suggest that species, site characteristics,
problem burden, diameter and geography may all contain useful predictive
information. These are associations only. They may also reflect differences
in species distribution, neighbourhood conditions, assessment behaviour or
other unmeasured factors.

# Initial Modelling Plan

A future modelling phase should compare an interpretable baseline against more
flexible non-linear classifiers. Suitable candidates include multinomial
logistic regression and tree-based ensembles. Preprocessing should be learned
inside the resampling workflow to prevent data leakage.

Because the outcome is imbalanced, evaluation should not focus on overall
accuracy alone. Macro-averaged precision, recall and F1, class-specific
sensitivity/recall, balanced accuracy and the confusion matrix are appropriate
complements. The Poor class is particularly important because it is both rare
and practically relevant.

Predictors considered from the current EDA include tree diameter, species,
stewardship, guards, sidewalk condition, problem burden, borough and spatial
coordinates. Variables that directly encode or leak the target should be
excluded. User type should be assessed cautiously because it may capture
measurement-process differences rather than underlying tree condition.

# Limitations and Responsible Interpretation

The census is observational and historical, so any model trained on it would
learn patterns from the 2015--2016 data collection process rather than provide
a timeless measure of tree health. The response is perceived health rather
than a clinical diagnosis, and potential measurement differences among census
participants should be considered.

A classification model should therefore be used only as a decision-support or
screening tool. Trees predicted to be at elevated risk should still receive
professional inspection before maintenance or removal decisions are made.

# References

New York City Department of Parks and Recreation. (2016). *2015 TreesCount!
Street Tree Census*. NYC Parks.

New York City Department of Parks and Recreation. (2024). *2015 Street Tree
Census - Tree Data*. NYC Open Data.
